Variance reduction for MCMC methods via martingale representations

### 2020-02-19 aap/Revision/vr-mcmc-revision-new.tex
update

\newcommand{\eric}[1]{\todo[color=purple!10, author=Eric]{#1}}
\newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
\newcommand{\denis}[1]{\todo[color=green!10, author=Denis]{#1}}
\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
\newcommand{\sergey}[1]{\todo[color=red!10, author=Sergey]{#1}}
\newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}

@@

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$. Then we can define

@@

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
\label{eq:qpl}
Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X_{p}),
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 1,\)} \eric{I do not understand this}  the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X_{N+l-1})]
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma=\gamma_1=\gamma_2=\ldots.$
\tcr{We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$.} \eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}
By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L_U < \infty$ such that
|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \rset^d.
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds
\ps{D^2U(x)}{x} \geq (m_U/2)\|x\|^2.
\end{description}
First we analyse the biases of the estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
\paragraph{Squared bias:} Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\mathsf E\left[M_{K,n}^N(f)\right]=0,
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$

@@

\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(4 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{4\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.

@@

@@

where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}

>>>>

\newcommand{\eric}[1]{\todo[color=purple!10, author=Eric]{#1}}
\newcommand{\erici}[1]{\todo[inline,color=purple!10, author=Eric]{#1}}
\newcommand{\denis}[1]{\todo[color=green!10, author=Denis]{#1}}
\newcommand{\denisi}[1]{\todo[inline,color=green!10, author=Denis]{#1}}
\newcommand{\sergey}[1]{\todo[color=blue!10, author=Sergey]{#1}}
\newcommand{\sergeyi}[1]{\todo[inline,color=red!10, author=Sergey]{#1}}

@@

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) We shall use the notations $\nset=\{1,2,\ldots\}$ and $\nset_0=\mathbb N\cup\{0\}$. We denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration generated by independent random variables $(\xi_p)_{p \in \mathbb{N}^*}$ with the convention $\mathcal{G}_0=\mathrm{triv}$. For $l \le p$ and $x \in \rset^d$, put for $x,y \in \rset^d$ functions $\Phi_k(x,y)=x-\gamma_k\mu(x)+y$. Then we can define

@@

From  numerical point of view another representation of the coefficients \(a_{p,l,k}\)  turns out to be more useful.
\begin{prop}
The coefficients \(a_{p,l,k}\) in \eqref{eq:coeff_mart}  can be alternatively represented as
\begin{eqnarray*}
a_{p,l,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{p,l}\left(\Phi_l(x,\xi)\right)\right]
\end{eqnarray*}
with \(Q_{p,l}(x)=\mathsf{E}\left[f(X^x_{l,p})\right],\) \(p\geq l.\)
%where   a r.v. \(\xi\) independent of \(\mathcal{G}_{l-1}.\)
The functions \((Q_{p,l})_{l=0}^p\)  can be computed by the backward recurrence: $Q_{p,p}(x)=f(x)$ and for $l \in \{0,\ldots,p-1\}$
\begin{eqnarray}
\label{eq:qpl}
Q_{p,l}(x)=\mathsf{E}\left[Q_{p,l+1}(X^x_{l+1})\right].
%=\mathsf{E}[Q_{p,l+1}(\Phi_{l+1}(x,\xi))]
\end{eqnarray}
In the case $\Phi_l=\Phi$ for all $l\geq 1,$ we have
\begin{equation}
\label{eq:definition-bar-a}
\bar a_{r,k}(x)=\mathsf{E}\left[\phi_k\left(\xi\right)Q_{r}\left(\Phi(x,\xi)\right)\right]
\end{equation}
with  \(Q_{r}(x)=\mathsf{E}\left[f(X^x_{r})\right],\) $r\in \mathbb{N}.$
\end{prop}
\par
Next we show how  the representation \eqref{eq:mart_repr} can be used to  reduce the variance of MCMC algorithms.  Consider the case of time homogeneous transition kernels and introduce a weighted average estimator $\pi_{n}^{N}(f)$ of the form
\begin{equation}\label{eq:29032018a2}
\pi_{n}^{N}(f)=\frac{1}{n}\sum_{p=N+1}^{N+n} f(X_{p}),
\end{equation}
where $N\in\mathbb N_0$ is the length of the burn-in period and $n\in\mathbb N$
the number of effective samples.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{eqnarray}
\nonumber
M_{K,n}^N(f) &=&\frac{1}{n}\sum_{p=N+1}^{N+n}\left[\sum_{k=1}^{K}\sum_{l=N+1}^{p} \bar a_{p-l,k}(X_{l-1})\phi_k(\xi_{l})\right]
\\
&=& \sum_{k=1}^{K}\sum_{l=N+1}^{N+n}\left(1+\frac{N-l}{n}\right)A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right),
\label{eq:29032018a5}
\end{eqnarray}
where
\begin{equation}
\label{eq:definition-A-s,k}
A_{s,k}(x)=\frac{1}{s}\sum_{r=1}^{s}\bar a_{r,k}(x).
\end{equation}
Since \(X_{l-1}\) is independent of \(\xi_{l}\) and \tcr{\(\mathsf{E}[\phi_k(\xi_{l})]=0,\) \(k\neq 0\)}, we obtain 
\[\mathsf{E}[A_{N+n-l,k}(X_{l-1})\phi_{k}\left(\xi_{l}\right)] = \mathsf{E}[A_{N+n-l,k}(X_{l-1})\mathsf{E}[\left.\phi_{k}\left(\xi_{l}\right)\right|\mathcal{G}_{l-1}]] = 0\]
\eric{I do not understand this} and hence the r.v.  \(M_{K,n}^N(f)\) has zero mean and can be viewed as a control variate.
The representation \eqref{eq:coeff_mart} suggests that the variance of  the  variance-reduced estimator
\begin{equation}
\label{eq:29032018a3}
\pi_{K,n}^N(f)=\pi_n^N(f)-M_{K,n}^N(f)
\end{equation}
should be small for \(K\) large enough. Indeed, since $\PE[ \phi_k(\xi_l) \phi_{k'}(\xi_l)]=0$ if $k \ne k'$, we obtain
\begin{equation}
\label{eq:expression-variance}
\PVar[\pi_{K,n}^N(f)]\leq\sum_{k=K+1}^{\infty}\sum_{l=1}^{n}\PE[A^2_{n-l,k}(X_{N+l-1})]
\end{equation}
and $\mathsf{Var}[\pi_{K,n}^N(f)]$ is small provided that the coefficients $A_{s,k}$ decay fast enough as $k\to \infty.$
In the next section   we provide a detailed   analysis of  $\mathsf{Var}[\pi_{K,n}^N(f)]$  for the  ULA  algorithm (see Example~\ref{exam:langevin-algorithm}).
\section{Analysis of variance reduced ULA}
\label{sec:ula_analysis}
In this section we perform the convergence analysis of the ULA algorithm. For the sake of clarity and notational simplicity we restrict our attention to the constant time step, that is, we take $\gamma_k =\gamma$ for any $k \in \nset$.

\sergey{N and N0 were defined here before}
\eric{these notations are used before; it is best to put all the notations in a short paragraph just after the introduction}

By $H_k$, $k\in\mathbb N_0$,
we denote the normalized Hermite polynomial on $\mathbb R$, that is,
$$
H_k(x)=\frac{(-1)^k}{\sqrt{k!}}\rme^{x^2/2}\frac{\partial^k}{\partial x^k}\rme^{-x^2/2},
\quad x\in\mathbb R.
$$
For a multi-index $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
$\mathbf{H}_\mathbf{k}$ denotes the normalized Hermite polynomial on $\mathbb R^d$, that is,
$$
\mathbf{H}_\mathbf{k}(\mathbf{x})=\prod_{i=1}^d H_{k_i}(x_i),\quad \mathbf{x}=(x_i)\in\mathbb R^d.
$$
In what follows, we also use the notation
$|\mathbf{k}|=\sum_{i=1}^d k_i$ for $\mathbf{k}=(k_i)\in\mathbb N_0^d$,
and we set $\mathcal G_p=\sigma(Z_1,\ldots,Z_p)$, $p\in\mathbb N$, and $\mathcal G_0=\mathrm{triv}$.
Given $N$ and $n$ as above, for $K\in\mathbb N$, denote
\begin{align}
M_{K,n}^N(f) & =\sum_{\mathbf k\colon 0<\|\mathbf{k}\|\le K}\sum_{l=1}^{n}\Bigl(1-\frac{l}{n}\Bigr)A_{n-l,\mathbf{k}}(X_{N+l-1})\mathbf{H}_\mathbf{k}(Z_{N+l})
\label{eq:29032018a5}
\notag
\end{align}
with \(\|\mathbf{k}\|=\max_{i} k_i.\)
For an estimator $\rho(f)\in\{\pi_n^N(f),\pi_{K,n}^N(f)\}$
of $\pi(f)$ (see \eqref{eq:29032018a2} and \eqref{eq:29032018a3}), we shall be interested in the Mean Squared Error (MSE),
which can be decomposed as the sum of the squared  bias
and the  variance:
\begin{equation}
\label{eq:29032018a4}
\mathrm{MSE}\left[\rho(f)\right] = \PE\left[\left\{\rho(f)-\pi(f)\right\}^2\right]
=\left\{\PE[\rho(f)]-\pi(f)\right\}^2 +\PVar[\rho(f)].
\end{equation}
Our analysis is carried out under the following two assumptions:
\begin{description}
\item[(H1)][\textbf{Lipschitz continuity}] The potential $U$ is differentiable and $\nabla U$ is Lipschitz, that is, there exists $L_U < \infty$ such that
\begin{eqnarray*}
|\nabla U(x)-\nabla U(y)|\leq L_U|x-y|, \quad x,y\in \rset^d.
\end{eqnarray*}
\item[(H2)][\textbf{Convexity outside a ball}] There exist $K_U>0,$ $M_U>0$ and $m_U>0$ such that for any $x\not\in B_{K_U}(0)$  it holds
\begin{equation*}
\ps{D^2U(x)}{x} \geq (m_U/2)\|x\|^2.
\end{equation*}
\end{description}
First we analyse the biases of the estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f).$
\paragraph{Squared bias:} Due to the martingale transform structure
of $M_{K,n}^N(f)$,
we have
\[
\mathsf E\left[M_{K,n}^N(f)\right]=0,
\]
Hence both estimators
$\pi_n^N(f)$ and $\pi_{K,n}^N(f)$ have the same conditional bias.
Notice that this remains true also when we substitute
the coefficients $a_{p,l,\mathbf{k}}$ in~\eqref{eq:29032018a5}
with some independent approximations $\widehat a_{p,l,\mathbf{k}}.$

@@

\kerULA_{\gamma}(x,A)=\int_{A} \frac{1}{(2 \pi \gamma)^{d/2}} \text{ exp} \left\{ -\frac{1}{2\gamma} \left\| y - x + \gamma \mu(x)\right\|^2\right\} \rmd y.

@@

\sergey{Changed notation here, put variance of the noise equal to $\gamma$}

@@

%where \tcr{\(\lesssim\) stands for inequality up to a constant not depending  on $\gamma,$ \(n\) and \(N.\)}
\sergey{removed repeating explanation of $\lesssim$}
